---
title: Pairwise Comparisons 
author: AJ Fagan
date: 30 July, 2024
---

```{r chunk-setup, echo=F}
knitr::opts_chunk$set(echo=F, message=F)
setwd("/mnt/hdd/AJNewton/rnaseq-2023/pairwise-comp/")
```

```{r packages}
library(dplyr)
library(DESeq2)
```

```{r data-loading}
df.raw <- read.csv("../base-data.txt")
df.raw <- df.raw %>%
  mutate(cts = exp(value) - 1) %>%
  dplyr::rename(logcts = value)
```

```{r deseq-fitting}
cts <- df.raw %>%
  mutate(cts = round(cts)) %>%
  tidyr::pivot_wider(
    id_cols = "gene",
    names_from = "group",
    values_from = "cts"
  ) %>%
  data.frame(row.names = "gene")
coldata <- data.frame(
  treatment = c(rep("DMSO", 9), rep("IC50", 9), rep("IC75", 9)),
  time = rep(c(1,1,1,24,24,24,8,8,8), 3),
  group = rep(c("b", "c", "a"), 9),
  batch = rep(c("May", "May", "March"), 9)
)
coldata$time <- factor(coldata$time, levels = c(1,8,24))
coldata$treatment <- factor(coldata$treatment, levels = c("DMSO", "IC50", "IC75"))
coldata$group <- factor(coldata$group, levels = c("a", "b", "c"))
coldata$batch <- factor(coldata$batch, levels = c("March", "May"))
# coldata = t(as.matrix(coldata))

dds <- DESeqDataSetFromMatrix(
  countData = cts,
  colData = coldata,
  design = ~ treatment * time + group
)
dds <- DESeq(dds, reduced = ~ time + group, test = "LRT")
```


# Overall $p$-values - LRT

Below are the results from the DESeq analysis, for the `r n=10; n` genes with the lowest adjusted $p$-value. 
Note that, while the log2FoldChange column indicates the estimated difference in expression between 45uM and DMSO at 24 hours, the pvalue and padj columns are *independent* of time and treatment.
Small $p$-values should be interpreted as

> There is enough evidence to suggest that the observed effect of the treatment, as it interacts with a time effect, is not due to random chance. Thus, we conclude that treatment with atovaquone has an effect on the expression of this gene for at least one of the time points present.

***NOT*** as

> ~~There is enough evidence to suggest that the observed expression of this gene in the presence of 45 uM Atovaquone at 24 hours compared to a control at the same time point, is not due to random chance. Thus, we conclude that treatment of 45 uM atovaquone has an effect on the expression of this gene at the 24 hour time point.~~

This is because we used a Likelihood Ratio-Test (LRT), which tests simply (and naively stated) if the estimated parameters are more likely to be true than the null parameters.

```{r deseq-results}
as.data.frame(results(dds, contrast = c(
                        0, # Intercept
                        0, # IC50
                        1, # IC75
                        0, # Time 8
                        0, # Time 24
                        0, # Batch b
                        0, # Batch c
                        0, # IC50 x Time 8
                        0, # IC75 x Time 8
                        0, # IC50 x Time 24
                        1  # IC75 x Time 24
                        ))) %>%
  arrange(padj) %>%
  select(log2FoldChange, pvalue, padj) %>%
  head(n) %>%
  mutate_if(is.numeric, ~ as.character(signif(., 3))) %>%
  knitr::kable()
```


# Groupwise $p$-values

## Concerns


If we want $p$-values corresponding to pairwise comparisons, (statistical mumbo-jumbo incoming, don't worry if you don't get most of it, I'll try to highlight the big picture ideas) we need to utilize a Wald test with contrasts.
While the Wald test is capable of giving us $p$-values for DE between different treatment/time groups, ***the results of the Wald test are less reliable than those of the LRT***.
This is especially pertinent due to the ***small sample size*** for each individual gene.

Notably, the Wald test ***struggles when all values being compared are identical***.
For example, if, for a gene $G$, we observed $0$ instances of $G$ in any of the samples at 24 hours for either DMSO or 45uM Atovaquone. 
The Wald test requires estimating the standard error of the point-estimate, and then constructs a test-statistic by dividing the point-estimate by the standard error. 
But, in this instance, the estimated standard error would be 0, causing a malformed test-statistic.
The LRT, on the other hand, handles this case fine, and will simply determine the $p$-value is $1.0$.
As such, ***this issue has the biggest effect on genes that are not actually DE***.
However, ***it can still present problems in less extreme cases, leading to an inflated type-I error rate***.

Other problems plague the Wald test, however. 
The ***Hauck-Donner Effect*** is another one that should be mentioned. 
It states that, in many cases, the standard error grows faster than the test-statistic.
This could result in **genes with truly high DE point-estimates _being falsely deemed insignificant_**.
Unfortunately, this result, unlike the previous, is more likely to effect the genes that are actually DE, than those that are not. 
Close attention will be needed to watch for this effect.

Finally, since we are using *multiple comparisons*, further $p$-value corrections will be needed to try to preserve our type-$I$ error rate. 
In particular, I'm going to apply a Bonferonni Correction to each comparison.
Since I'm performing 6 comparisons (comparing each pair of groups ($\begin{pmatrix}3\\2\end{pmatrix} = 3$) at both 8 and 24 hours), I will multiply each of the FDR-adjusted $p$-values by 6.


```{r contrast-definition}
contrasts <- list(
  h24_45vD = c(
    0, # Intercept
    0, # IC50
    1, # IC75
    0, # Time 8
    0, # Time 24
    0, # Batch b
    0, # Batch c
    0, # IC50 x Time 8
    0, # IC75 x Time 8
    0, # IC50 x Time 24
    1  # IC75 x Time 24
  ),
  h24_45v30 = c(
    0, # Intercept
    -1, # IC50
    1, # IC75
    0, # Time 8
    0, # Time 24
    0, # Batch b
    0, # Batch c
    0, # IC50 x Time 8
    0, # IC75 x Time 8
    -1, # IC50 x Time 24
    1  # IC75 x Time 24
  ),
  h24_30vD = c(
    0, # Intercept
    1, # IC50
    0, # IC75
    0, # Time 8
    0, # Time 24
    0, # Batch b
    0, # Batch c
    0, # IC50 x Time 8
    0, # IC75 x Time 8
    1, # IC50 x Time 24
    0  # IC75 x Time 24
  ),
  h8_45vD = c(
    0, # Intercept
    0, # IC50
    1, # IC75
    0, # Time 8
    0, # Time 24
    0, # Batch b
    0, # Batch c
    0, # IC50 x Time 8
    1, # IC75 x Time 8
    0, # IC50 x Time 24
    0  # IC75 x Time 24
  ),
  h8_45v30 = c(
    0, # Intercept
    -1, # IC50
    1, # IC75
    0, # Time 8
    0, # Time 24
    0, # Batch b
    0, # Batch c
    -1, # IC50 x Time 8
    1, # IC75 x Time 8
    0, # IC50 x Time 24
    0  # IC75 x Time 24
  ),
  h8_30vD = c(
    0, # Intercept
    1, # IC50
    0, # IC75
    0, # Time 8
    0, # Time 24
    0, # Batch b
    0, # Batch c
    1, # IC50 x Time 8
    0, # IC75 x Time 8
    0, # IC50 x Time 24
    0  # IC75 x Time 24
  )
)

save.gene.results <- function(df, name, dir = "output/", extension = ".csv") {
  df %>%
    tibble::rownames_to_column("gene") %>%
    write.table(
      ., 
      paste(
        dir, 
        name,
        extension,
        sep = ""
      ),
      row.names = F
    )
  df
}
show.gene.results <- function(
    dds, 
    contrast, 
    n = 10, 
    columns = c("baseMean", "log2FoldChange", "lfcSE", "stat", "pvalue", "padj"),
    save = T, dir = "output/", extension = ".csv") {
  results(dds, test = "Wald", contrast = contrast[[1]]) %>%
    as.data.frame() %>%
    arrange(padj) %>%
    mutate(padj = 6*padj) %>%
    save.gene.results(names(contrast), dir = dir, extension = extension) %>%
    head(n)
}
```


## 24 hours

### 45 uM vs DMSO

```{r}
show.gene.results(dds, contrasts[1])
```


### 45 uM vs 30 uM

```{r}
show.gene.results(dds, contrasts[2])
```

### 30 uM vs DMSO

```{r}
show.gene.results(dds, contrasts[3])
```

## 8 hours

### 45 uM vs DMSO

```{r}
show.gene.results(dds, contrasts[4])
```


### 45 uM vs 30 uM

```{r}
show.gene.results(dds, contrasts[5])
```

### 30 uM vs DMSO

```{r}
show.gene.results(dds, contrasts[6])
```




