---
title: "Follow-up on Meeting from 20 August, 2024"
output: html_notebook
author: "AJ Fagan"
date: "23 Aug, 2024"
---


```{r chunk-setup, echo=F}
knitr::opts_chunk$set(echo=F, message=F)
n=10
```

```{r packages, include=F, echo=F}
library(dplyr)
library(ggplot2)
library(DESeq2)
library(pheatmap)
library(tidyverse)
library(org.Hs.eg.db)
library(allez)
library(pathfindR)
```


```{r data-loading}
df.raw <- read.csv("../../base-data.txt")
df.raw <- df.raw %>%
  mutate(cts = exp(value) - 1) %>%
  dplyr::rename(logcts = value)
```

```{r deseq-fitting}
cts <- df.raw %>%
  mutate(cts = round(cts)) %>%
  tidyr::pivot_wider(
    id_cols = "gene",
    names_from = "group",
    values_from = "cts"
  ) %>%
  data.frame(row.names = "gene")
coldata <- data.frame(
  treatment = c(rep("DMSO", 9), rep("IC50", 9), rep("IC75", 9)),
  time = rep(c(1,1,1,24,24,24,8,8,8), 3),
  group = rep(c("b", "c", "a"), 9),
  batch = rep(c("May", "May", "March"), 9)
)
coldata$time <- factor(coldata$time, levels = c(1,8,24))
coldata$treatment <- factor(coldata$treatment, levels = c("DMSO", "IC50", "IC75"))
coldata$group <- factor(coldata$group, levels = c("a", "b", "c"))
coldata$batch <- factor(coldata$batch, levels = c("March", "May"))
# coldata = t(as.matrix(coldata))

dds <- DESeqDataSetFromMatrix(
  countData = cts,
  colData = coldata,
  design = ~ treatment * time + group
)
dds <- DESeq(dds, reduced = ~ time + group, test = "LRT")
```

```{r utility-functions}
# Format plot title
wrapper <- function(x, ...)
{
  paste(strwrap(x, ...), collapse = "\n")
}

```

# Preface


I'm copy-pasting much from the previous report to save time. 

> I will put new information/changes in block quotes like this. 

# Intro

Hey Mayra,



Below are (at least several of) the plots you asked for.
Let me know if I missed any.

For each type of plot, I included several replicates for varying number of genes to help you pick a number to display that you like. 

# Heatmap of DE between each pair of treatment groups at each time

## Ordered by $p$-value

For each pair of treatment groups, and time combination, shows the estimated log2-fold change in expression. 
~~For example, for the green-time, red-treatment, shows how much more (or less) highly expressed each gene was at 24 hours in the 45$\mu$M group compared to the 30$\mu$M group.~~

> Removed 45vs30 Comparison. Now the green-time, purple-treatment column shows how much more (or less) highly expressed each gene was at 24 hours in the 45$\mu$M group compared to the DMSO group.

```{r de-over-time}
# 75 v DMSO @ 24
de.time <- data.frame(
  fc75.D.24 = results(dds, contrast = c(
                        0, # Intercept
                        0, # IC50
                        1, # IC75
                        0, # Time 8
                        0, # Time 24
                        0, # Batch b
                        0, # Batch c
                        0, # IC50 x Time 8
                        0, # IC75 x Time 8
                        0, # IC50 x Time 24
                        1  # IC75 x Time 24
                        ))$log2FoldChange
)

# 50 v DMSO @ 24
de.time$fc50.D.24 <- results(dds, contrast = c(
                        0, # Intercept
                        1, # IC50
                        0, # IC75
                        0, # Time 8
                        0, # Time 24
                        0, # Batch b
                        0, # Batch c
                        0, # IC50 x Time 8
                        0, # IC75 x Time 8
                        1, # IC50 x Time 24
                        0  # IC75 x Time 24
                        ))$log2FoldChange

# 75 v DMSO @ 8
de.time$fc75.D.8 <- results(dds, contrast = c(
                        0, # Intercept
                        0, # IC50
                        1, # IC75
                        0, # Time 8
                        0, # Time 24
                        0, # Batch b
                        0, # Batch c
                        0, # IC50 x Time 8
                        1, # IC75 x Time 8
                        0, # IC50 x Time 24
                        0  # IC75 x Time 24
                        ))$log2FoldChange
# 50 v DMSO @ 8
de.time$fc50.D.8 <- results(dds, contrast = c(
                        0, # Intercept
                        1, # IC50
                        0, # IC75
                        0, # Time 8
                        0, # Time 24
                        0, # Batch b
                        0, # Batch c
                        1, # IC50 x Time 8
                        0, # IC75 x Time 8
                        0, # IC50 x Time 24
                        0  # IC75 x Time 24
                        ))$log2FoldChange

# 75 v DMSO @ 1
de.time$fc75.D.1 <- results(dds, contrast = c(
                        0, # Intercept
                        0, # IC50
                        1, # IC75
                        0, # Time 8
                        0, # Time 24
                        0, # Batch b
                        0, # Batch c
                        0, # IC50 x Time 8
                        0, # IC75 x Time 8
                        0, # IC50 x Time 24
                        0  # IC75 x Time 24
                        ))$log2FoldChange
# 50 v DMSO @ 1
de.time$fc50.D.1 <- results(dds, contrast = c(
                        0, # Intercept
                        1, # IC50
                        0, # IC75
                        0, # Time 8
                        0, # Time 24
                        0, # Batch b
                        0, # Batch c
                        0, # IC50 x Time 8
                        0, # IC75 x Time 8
                        0, # IC50 x Time 24
                        0  # IC75 x Time 24
                        ))$log2FoldChange

rownames(de.time) <- rownames(results(dds))
de.time <- de.time[complete.cases(de.time),]

gene_ord_all <- list(
  "p" = rownames(results(dds)[order(results(dds)$padj, decreasing = F),]),
  "up" = rownames(results(dds)[order(results(dds)$log2FoldChange, decreasing = T),]),
  "down" = rownames(results(dds)[order(results(dds)$log2FoldChange, decreasing = F),]),
  'any' = rownames(results(dds)[order(abs(results(dds)$log2FoldChange), decreasing = T),])
)
gene_ord <- sapply(gene_ord_all, function(x) {
  x <- x[!is.na(results(dds)[x,"padj"]) & (results(dds)[x,"padj"] < 0.05)]
}, USE.NAMES = T, simplify = F)
gene_ord[['p']] <- gene_ord_all[['p']]

de.time <- de.time[complete.cases(de.time),]


# Plot a heatmap of diff. ex
treatment.pairs <- c(
"45uM v DMSO",
"30uM v DMSO"
)
de.coldata <- data.frame(
  treatment = rep(c("45uM v DMSO", "30uM v DMSO"), 3),
  time = rep(c(24, 8, 1), each = 2)
)
rownames(de.coldata) <- colnames(de.time_p)
de.heatmap <- function(
  n,
  x = de.time,
  treatment_pairs = treatment.pairs,
  times = c(1, 8, 24),
  col_data = de.coldata,
  title = function(n, order_by) {
    switch(order_by,
                  p = paste(
                    "Estimated log2-fold change for each treatment group at each time for the",
                    n,
                    "genes with the smallest p-value."
                  ),
                  up = paste(
                    "Estimated log2-fold change for each treatment group at each time for the",
                    n,
                    "genes with the greatest log fold-change with p<0.05."
                  ),
                  down = paste(
                    "Estimated log2-fold change for each treatment group at each time for the",
                    n,
                    "genes with the least log fold-change with p<0.05."
                  ),
                  any = paste(
                    "Estimated log2-fold change for each treatment group at each time for the",
                    n,
                    "genes with the greatest absolute log fold-change with p<0.05."
                  )
    )
  },
  title_wrapper = 40,
  order_by = c("p", "up", "down", "any")
) {
  order_by <- match.arg(order_by)
  
  x <- x[gene_ord[[order_by]],]
  col_data <- col_data %>%
    filter(time %in% times) %>%
    mutate(time = as.factor(time)) %>%
    filter(treatment %in% treatment_pairs) %>%
    mutate(treatment = factor(treatment, ordered = T)) %>%
    #arrange(treatment) %>%
    dplyr::rename("Time (Hours)" = time) %>%
    dplyr::rename("Treatment Pair" = treatment)
  if (length(times) <= 1) {
    col_data <- col_data %>% dplyr::select(-c("Time (Hours)"))
  }
  x <- x[,colnames(x) %in% rownames(col_data)]
  pheatmap(
    as.matrix(x[1:n,]),
    annotation_col = col_data,
    cluster_rows = F, cluster_cols = F, show_colnames = F,
    legend_labels = c("Time (Hours)", "Treatment Pair"),
    main = wrapper(title(n, order_by), title_wrapper),
    legend = T
  )
}
```

```{r}
de.heatmap(5)
de.heatmap(10)
de.heatmap(20)
de.heatmap(25)
de.heatmap(30)
de.heatmap(50)
```


## Ordered by estimated effect size, restricted to FDR-controlled $p$-value < 0.05

> Now I show essentially the same images, but now we sort them by estimated effect size at 24 hours between 45$\mu$M and DMSO. For brevity, I include only one image for each, rather than for a range of #genes, but can trivially include more, if desired.

### Ordered by decreasing log fold-change (i.e. up-regulated)

```{r}
de.heatmap(25, order_by = "up")
```



### Ordered by increasing log fold-change (i.e. down-regulated)

```{r}
de.heatmap(25, order_by="down")
```


### Ordered by decreasing abs(log fold-change) (i.e. effect far from zero)

```{r}
de.heatmap(25, order_by="any")
```



# Barplot of DE between each pair of treatment groups at each time

## Ordered by FDR-adjusted $p$


This contains pretty much the same information as above, but in a barplot.
It may be slightly easier to glean info from this, but fewer genes will fit while still looking nice.

> I also made the same changes here that I did above.

```{r}
genes.ordered <- factor(rownames(de.time), ordered=T)
de.time2 <- de.time
de.time2$gene <- rownames(de.time2)
de.time2 <- de.time2 %>%
  pivot_longer(-c(gene)) %>%
  mutate(treatment = de.coldata[name, "treatment"]) %>%
  mutate(time = as.factor(de.coldata[name,"time"])) %>%
  dplyr::rename(log2FoldChange = value)

de.barplot <- function(
    n, 
    genes = factor(rownames(de.time)),
    x = de.time2, 
    title = function(n, order_by) {
      switch(order_by,
                  p = paste(
                    "Bar plot of estimated log2-fold change for the",
                    n,
                    "genes with the smallest p-value."
                  ),
                  up = paste(
                    "Bar plot of estimated log2-fold change for the",
                    n,
                    "genes with the greatest log fold-change with p<0.05."
                  ),
                  down = paste(
                    "Bar plot of estimated log2-fold change for the",
                    n,
                    "genes with the least log fold-change with p<0.05."
                  ),
                  any = paste(
                    "Bar plot of estimated log2-fold change for the",
                    n,
                    "genes with the greatest absolute log fold-change with p<0.05."
                  )
      )
    },
    order_by = c("p", "up", "down", "any")
) {
  order_by <- match.arg(order_by)
  #print(head(de.time2 %>% filter(gene %in% gene_ord[[order_by]][1:n])))
  #print(length(unique((de.time2 %>% filter(gene %in% gene_ord[[order_by]][1:n]))$gene)))
  
  #print(head(de.time2 %>% filter(gene %in% gene_ord[[order_by]][1:n]) %>% filter(gene %in% genes)))
  ggplot(
    de.time2 %>% filter(gene %in% gene_ord[[order_by]][gene_ord[[order_by]] %in% genes][1:n]),
    aes(
      y = gene,
      x = log2FoldChange,
      fill = time
    )
  ) +
    geom_col(position= position_dodge(width = 0.9)) +
    facet_wrap(~treatment) +
    xlab("log2-Fold Change") +
    ylab("Gene") +
    ggtitle(wrapper(title(n, order_by))) +
    ylim(gene_ord[[order_by]][gene_ord[[order_by]] %in% genes][n:1])
}
de.barplot(5)
de.barplot(10)
de.barplot(20)
de.barplot(30)
de.barplot(50)
```


## Ordered by estimated effect size, restricted to FDR-controlled $p$-value < 0.05

### Ordered by decreasing log fold-change (i.e. up-regulated)

```{r}
de.barplot(30, order_by = 'up')
```


### Ordered by increasing log fold-change (i.e. down-regulated)

```{r}
de.barplot(30, order_by = 'down')
```


### Ordered by decreasing absolute log fold-change (i.e. effect far from zero)

```{r}
de.barplot(30, order_by = 'any')
```




# Heatmap of log-counts

> I am not including this currently. Can do so upon request.


# Bar plot for ERAD genes

## Ordered by increasing $p$

This contains essentially the same info as the barplots above, but specialized to ERAD (GO:0036503) genes. 

I could easily also present any of the above graphics for any given geneset. 

> Only 28 genes in ERAD had p < 0.05

```{r}
erad.genes <- (as.data.frame(org.Hs.egALIAS2EG) %>%
  filter(gene_id %in% as.data.frame(org.Hs.egGO2ALLEGS["GO:0036503"])$gene_id) %>%
  filter(alias_symbol %in% rownames(results(dds))) %>%
  filter(alias_symbol %in% df.raw$gene))$alias_symbol
'VCP' %in% erad.genes

de.barplot(5, genes=erad.genes, title = function(n, order_by) {paste(
  "Bar plot of estimated log2-fold change for the",
  n,
  "genes in GO:0036503 with the lowest p-value."
)})
de.barplot(10, genes=erad.genes, title = function(n, order_by) {paste(
  "Bar plot of estimated log2-fold change for the",
  n,
  "genes in GO:0036503 with the lowest p-value."
)})
de.barplot(20, erad.genes, title = function(n, order_by) {paste(
  "Bar plot of estimated log2-fold change for the",
  n,
  "genes in GO:0036503 with the lowest p-value."
)})
de.barplot(30, erad.genes, title = function(n, order_by) {paste(
  "Bar plot of estimated log2-fold change for the",
  n,
  "genes in GO:0036503 with the lowest p-value."
)})
de.barplot(50, erad.genes, title = function(n, order_by) {paste(
  "Bar plot of estimated log2-fold change for the",
  n,
  "genes in GO:0036503 with the lowest p-value."
)})
```


### Ordered by decreasing log fold-change (i.e. up-regulated)



```{r}
de.barplot(28, erad.genes, order_by = 'up', title = function(n, order_by) {paste(
  "Bar plot of estimated log2-fold change for the",
  n,
  "genes in GO:0036503 with the greatest log fold-change with p<0.05."
)})
```


### Ordered by increasing log fold-change (i.e. down-regulated)

```{r}
de.barplot(28, erad.genes, order_by = 'down', title = function(n, order_by) {paste(
  "Bar plot of estimated log2-fold change for the",
  n,
  "genes in GO:0036503 with the least log fold-change with p<0.05."
)})
```


### Ordered by decreasing absolute log fold-change (i.e. effect far from zero)

```{r}
de.barplot(28, erad.genes, order_by = 'up', title = function(n, order_by) {paste(
  "Bar plot of estimated log2-fold change for the",
  n,
  "genes in GO:0036503 with the greatest absolute log fold-change with p<0.05."
)})
```




# Oxidative stress barplots


## Ordered by increasing $p$


```{r}
erad.genes <- (as.data.frame(org.Hs.egALIAS2EG) %>%
  filter(gene_id %in% as.data.frame(org.Hs.egGO2ALLEGS["GO:0006979"])$gene_id) %>%
  filter(alias_symbol %in% rownames(results(dds))) %>%
  filter(alias_symbol %in% df.raw$gene))$alias_symbol

de.barplot(5, genes=erad.genes, title = function(n, order_by) {paste(
  "Bar plot of estimated log2-fold change for the",
  n,
  "genes in GO:0006979 with the lowest p-value."
)})
de.barplot(10, genes=erad.genes, title = function(n, order_by) {paste(
  "Bar plot of estimated log2-fold change for the",
  n,
  "genes in GO:0006979 with the lowest p-value."
)})
de.barplot(20, erad.genes, title = function(n, order_by) {paste(
  "Bar plot of estimated log2-fold change for the",
  n,
  "genes in GO:0006979 with the lowest p-value."
)})
de.barplot(30, erad.genes, title = function(n, order_by) {paste(
  "Bar plot of estimated log2-fold change for the",
  n,
  "genes in GO:0006979 with the lowest p-value."
)})
de.barplot(50, erad.genes, title = function(n, order_by) {paste(
  "Bar plot of estimated log2-fold change for the",
  n,
  "genes in GO:0006979 with the lowest p-value."
)})
```


### Ordered by decreasing log fold-change (i.e. up-regulated)



```{r}
de.barplot(30, erad.genes, order_by = 'up', title = function(n, order_by) {paste(
  "Bar plot of estimated log2-fold change for the",
  n,
  "genes in GO:0006979 with the greatest log fold-change with p<0.05."
)})
```


### Ordered by increasing log fold-change (i.e. down-regulated)

```{r}
de.barplot(30, erad.genes, order_by = 'down', title = function(n, order_by) {paste(
  "Bar plot of estimated log2-fold change for the",
  n,
  "genes in GO:0006979 with the least log fold-change with p<0.05."
)})
```


### Ordered by decreasing absolute log fold-change (i.e. effect far from zero)

```{r}
de.barplot(30, erad.genes, order_by = 'any', title = function(n, order_by) {paste(
  "Bar plot of estimated log2-fold change for the",
  n,
  "genes in GO:0006979 with the greatest absolute log fold-change with p<0.05."
)})
```

# Mitochondrial Fusion genes

## Ordered by increasing $p$


```{r}
erad.genes <- (as.data.frame(org.Hs.egALIAS2EG) %>%
  filter(gene_id %in% as.data.frame(org.Hs.egGO2ALLEGS["GO:0008053"])$gene_id) %>%
  filter(alias_symbol %in% rownames(results(dds))) %>%
  filter(alias_symbol %in% df.raw$gene))$alias_symbol

de.barplot(5, genes=erad.genes, title = function(n, order_by) {paste(
  "Bar plot of estimated log2-fold change for the",
  n,
  "genes in GO:0008053 with the lowest p-value."
)})
de.barplot(10, genes=erad.genes, title = function(n, order_by) {paste(
  "Bar plot of estimated log2-fold change for the",
  n,
  "genes in GO:0008053 with the lowest p-value."
)})
de.barplot(20, erad.genes, title = function(n, order_by) {paste(
  "Bar plot of estimated log2-fold change for the",
  n,
  "genes in GO:0008053 with the lowest p-value."
)})
de.barplot(30, erad.genes, title = function(n, order_by) {paste(
  "Bar plot of estimated log2-fold change for the",
  n,
  "genes in GO:0008053 with the lowest p-value."
)})
de.barplot(32, erad.genes, title = function(n, order_by) {paste(
  "Bar plot of estimated log2-fold change for the",
  n,
  "genes in GO:0008053 with the lowest p-value."
)})
```
> None had $p < 0.05$, so can't make the other plots


# Misc gene sets

> Found a bug in my previous code that removed ~6000 genes with $p\approx 1$ from the analysis. I also changed the type-I error rate to 0.05. GSEA results are slightly different as a result.

Below is a waterfall plot of the Allez-significant gene sets.
It shows, first, the gene-set that contains the most genes with adjusted $p < 0.05$, and plots that number of genes.
Then it, iteratively, selects the next best such set, but considering only those genes not contained in the previously selected sets.

Notably, Proteasomal Protein Catabolic Process contains the ERAD pathway as a child term, and Intrinsic Aptoptotic Signaling Pathway contains Intrinsic ... in Response to Oxidative Stress as child term.

```{r}
allez.dat <- as.numeric(!is.na(results(dds)$padj) & (results(dds)$padj < 0.05))
names(allez.dat) <- rownames(dds)
allez.out <- allez(allez.dat, "org.Hs.eg", idtype = "SYMBOL")
allezPlot(allez.out)
#allezTable(allez.out)
```



## pathfindR

I found this nifty tool that performs simultaneous network/gene set enrichment analysis in KEGG pathways.
I have not yet been able to look into the mathematics of what it does, so I can't speak to the meaningfulness of these results.
But it provides some nice visualization tools.

```{r}
path.out <- run_pathfindR(
  as.data.frame(results(dds)) %>% 
    mutate(gene = rownames(.)) %>% 
    dplyr::select(gene, log2FoldChange, padj) %>%
    mutate(padj = ifelse(is.na(padj), 1, padj))
)
as.data.frame(results(dds)) %>% 
    mutate(gene = rownames(.)) %>% 
    dplyr::select(log2FoldChange, padj) %>%
    mutate(padj = ifelse(is.na(padj), 1, padj))
```

See the term_visualizations folder for a real treat.
Those show the KEGG pathways, with DE genes highlighted green (up-regulated) or red (down-regulated).
As I mentioned earlier, I don't yet know the math behind how it got these values, so not publication ready yet, as far as I'm concerned.
But they at least provide a nice visualization to help elucidate what should be further explored. 

You can also see the kegg_pathways.csv file for information on each pathway.

A hierarchical clustering of the top-10 KEGG pathways is displayed below. 


```{r}
set.seed(42)
clu <- cluster_enriched_terms(path.out[1:10,], use_description = F, method="hierarchical")
```

The plot below shows a plot of the genes of the top 3 terms, along with teh direction of change. 
As you can see, these three terms share very few genes in common. 


```{r}
term_gene_heatmap(path.out, num_terms = 3, use_description = T)
```

And here is an attempt to plot a graph of the top pathways. 
It is very hard to interpret. 
I may be able to make it more legible, if this is of interest.

```{r}
term_gene_graph(path.out, num_terms = 3)
```


# Comparison Comparisons
```{r contrast-def}
contrasts <- list(
  h24_45vD = c(
    0, # Intercept
    0, # IC50
    1, # IC75
    0, # Time 8
    0, # Time 24
    0, # Batch b
    0, # Batch c
    0, # IC50 x Time 8
    0, # IC75 x Time 8
    0, # IC50 x Time 24
    1  # IC75 x Time 24
  ),
  h24_30vD = c(
    0, # Intercept
    1, # IC50
    0, # IC75
    0, # Time 8
    0, # Time 24
    0, # Batch b
    0, # Batch c
    0, # IC50 x Time 8
    0, # IC75 x Time 8
    1, # IC50 x Time 24
    0  # IC75 x Time 24
  ),
  h8_45vD = c(
    0, # Intercept
    0, # IC50
    1, # IC75
    0, # Time 8
    0, # Time 24
    0, # Batch b
    0, # Batch c
    0, # IC50 x Time 8
    1, # IC75 x Time 8
    0, # IC50 x Time 24
    0  # IC75 x Time 24
  ),
  h8_30vD = c(
    0, # Intercept
    1, # IC50
    0, # IC75
    0, # Time 8
    0, # Time 24
    0, # Batch b
    0, # Batch c
    1, # IC50 x Time 8
    0, # IC75 x Time 8
    0, # IC50 x Time 24
    0  # IC75 x Time 24
  )
)
```


```{r}
dd <- lapply(contrasts, function(x) {
  results(dds, test = "Wald", contrast = x) %>% 
    as.data.frame() %>% 
    tibble::rownames_to_column("gene") %>%
    mutate(sig = as.numeric(!is.na(padj) & padj < 0.01 / 4))
}) %>%
  bind_rows(.id = "comparison") %>%
  dplyr::select(comparison, gene, sig)
  #tidyr::pivot_wider(names_from = "gene", values_from = sig) %>%
  #data.frame(row.names = "comparison") %>%
  #as.matrix(dimnames = list(rownames(.), NULL)) %>%

compdata <- data.frame(
  time = rep(c("24 Hours", "8 Hours"), each = 2),
  d1 = rep(c("45 uM", "30 uM"), 2),
  d2 = rep(c("DMSO", "DMSO"), 2),
  row.names = unique(dd$comparison)
)

dd <- dd %>%
  {
    dd <- .
    ccs <- unique(dd$comparison)
    combs <- combn(ccs, 2, simplify = T)
    combs <- cbind(combs, rbind(combs[2,], combs[1,]), rbind(ccs, ccs))
    data.frame(
      c1 = combs[1,],
      c2 = combs[2,],
      n = apply(combs, 2, function(cc) {
        (dd %>%
          filter(comparison %in% cc) %>%
          group_by(gene) %>%
          summarize(b = sum(sig) == length(unique(cc))) %>%
          summarize(b = sum(b)))[1,1, 1]
      }),
      time1 = sapply(combs[1,], function(x) compdata[x, "time"]),
      time2 = sapply(combs[2,], function(x) compdata[x, "time"]),
      d11 = sapply(combs[1,], function(x) compdata[x, "d1"]),
      d12 = sapply(combs[1,], function(x) compdata[x, "d2"]),
      d21 = sapply(combs[2,], function(x) compdata[x, "d1"]),
      d22 = sapply(combs[2,], function(x) compdata[x, "d2"])
    )
  }

dd2 <- dd %>% 
  dplyr::select(-c(c1, c2)) %>%
  arrange(time2, d21, d22) %>%
  tidyr::pivot_wider(
    id_cols = c("time1", "d11", "d12"),
    names_from = c("time2", "d21", "d22"),
    values_from = n,
    values_fn = mean
  )# %>%
as.data.frame(dd2) %>%
  arrange(time1, d11, d12) %>%
  dplyr::select(-time1) %>%
  dplyr::rename("Treatment 1" = "d11") %>%
  dplyr::rename("Treatment 2" = "d12") %>%
  knitr::kable(align = 'llrrrrrr') %>% 
  #kableExtra::collapse_rows(1, valign = "top", latex_hline = "major") %>%
  #kableExtra::kbl(booktabs = T) %>%
  kableExtra::pack_rows(group_label = "Time", index = table(forcats::fct_inorder(as.character(dd2$time1)), useNA = "no")) %>%
  #kableExtra::pack_rows(group_label = "Treatment 1", index = table(forcats::fct_inorder(as.character(dd2$d11)), useNA = "no")) %>%
  kableExtra::header_separate(sep = "_")
dd
```


```{r}
dd <- lapply(contrasts, function(x) {
  results(dds, test = "Wald", contrast = x) %>% 
    as.data.frame() %>% 
    tibble::rownames_to_column("gene") %>%
    mutate(sig = as.numeric(!is.na(padj) & padj < 0.01 / 6))
}) %>%
  bind_rows(.id = "comparison") %>%
  dplyr::select(comparison, gene, sig)
  #tidyr::pivot_wider(names_from = "gene", values_from = sig) %>%
  #data.frame(row.names = "comparison") %>%
  #as.matrix(dimnames = list(rownames(.), NULL)) %>%

compdata <- data.frame(
  time = rep(c("24 Hours", "8 Hours"), each = 2),
  d1 = rep(c("45 uM", "30 uM"), 2),
  d2 = rep(c("DMSO", "DMSO"), 2),
  row.names = unique(dd$comparison)
)

dd2 <- dd %>%
  {
    dd <- .
    ccs <- unique(dd$comparison)
    dd2 <- data.frame(
      expand.grid(ccs, ccs, c(0,1))
    )
    colnames(dd2) <-  c('c1', 'c2', 'diff')
    dd2 <- dd2 %>% filter((c1 != c2) | (diff == 1))
    dd2 %>% 
      mutate(
        diff = as.numeric(diff),
        time1 = sapply(c1, function(x) compdata[x, "time"]),
        time2 = sapply(c2, function(x) compdata[x, "time"]),
        d11 = sapply(c1, function(x) compdata[x, "d1"]),
        d12 = sapply(c1, function(x) compdata[x, "d2"]),
        d21 = sapply(c2, function(x) compdata[x, "d1"]),
        d22 = sapply(c2, function(x) compdata[x, "d2"])
      ) 
    dd2$genes <- apply(dd2, MARGIN=1, function(x) {
      if (x['c1'] == x['c2']) {
        (dd %>%
          filter(
            ((comparison == x[['c1']]) & (as.numeric(sig) == 1))
          ) %>%
          group_by(gene) %>%
          summarize(n = n()) %>%
          filter(n > 0) %>%
          summarize(g = paste(gene, collapse = ";")))$g
      } else {
      #print(x)
        (dd %>%
          filter(
            ((comparison == x[['c1']]) & (as.numeric(sig) == 1)) | (
               (comparison == x[['c2']]) & (sig == 1-as.numeric(x[['diff']]))
            )
          ) %>%
          group_by(gene) %>%
          summarize(n = n()) %>%
          filter(n == 2) %>%
          summarize(g = paste(gene, collapse = ";")))$g
        
      }
    })
    dd2
  }
dd2$genes
dd2
mylist <- list()
out <- dd2 %>%
  mutate(c2 = ifelse(c1 == c2, "", as.character(c2))) %>%
  arrange(c1, c2, 1-diff) %>%
  mutate(c2 = apply(., 1, function(x) {
    if (x['diff'] == 1) {
      x['c2']
    } else {
      paste('-', x['c2'], sep = "")
    }
  })) %>%
  dplyr::select(c1, c2, genes) %>% 
  apply(., 1, function(x) {
    # ll <- list()
    # ll[[x[[2]]]] <- strsplit(x[[3]], ";")[[1]]
    # mylist[[x[[1]]]] <- ll
    strsplit(x[[3]], ";")
    #print(length(strsplit(x[['genes']], ";")[[1]]))
  }, simplify=F)
names(out)
names(out) <- apply(dd2, 1, function(x) {
  if (x['c1'] == x['c2']) {
    as.character(x['c1'])
  } else if (x['diff'] == 1) {
    paste(x['c1'], "and not", x['c2'])
  } else {
    paste(x['c1'], "and", x['c2'])
  }
})


outdf <- data.frame(a = out[[1]])
colnames(outdf) <- names(out[1])
for (i in 2:length(out)) {
  x <- out[[i]][[1]]
  x <- c(x, replicate(nrow(outdf) - length(x), ""))
  outdf[,names(out)[i]] <- x
}
dim(outdf)
head(outdf)
write.table(outdf, "gene-pairwise-significance.csv", sep = ",")
```


